---
title: "STA 523 Final Project"
author: "Dipesh Gautam, Radhika Anand, Haosheng Luo, Siyang Li"
date: "12/12/2014"
output: html_document
---

For this project, we have built a package called `concomp` to detect connected components and holes and their boundaries in an image or a logical matrix.

##Motivation
While working on 523 Homework-3, where we had to detect Manhattan precinct boundaries, we had data points in all of Manhattan apart from the Central Park area. While trying to detect the Central Park, we were looking for functions to detect holes and boundaries and hence decided to create a package with functions for the same.

##Functions
We have written the following three functions in our package:

1) `conLabel()` Assigns labels to connected components in a logical matrix

2) `conBoundaries()` Detects internal and external boundaries of connected components in a logical matrix

3) `imgMat()` Converts a PNG file to a logical matrix

##Methodology

####1) `conLabel(x,N=4)` Assigns labels to connected components in a logical matrix

#####Inputs

1) A logical matrix whose connected components are to be labeled

2) An integer to specify 4 or 8 connectivity

#####Output

Matrix with connected components detected and separately labeled

#####Algorithm

We have implemented a single-pass algorithm where we traverse the matrix row wise and label the connected components using 4 or 8-connectivity as specified by the user. In 4-connectivity, we look at the top, left, right and bottom neighbor of each cell and in 8-connectivity we check all 8 neighbors.

#####Steps

1) A connected-component matrix is initialized to size of the input matrix

2) A `mark` is initialized and incremented for every detected component in the matrix

3) A row-major scan is started for the entire matrix

4) If a component cell is detected, the following steps are repeated while (`index` !=0) 
+ Set the corresponding cell to 0 in the matrix
+ A vector `index` is updated with all the neighboring cells (4 or 8) of the currently set cells 
+ Only the unique cells are retained 
+ Set the cells indicated by `index` to `mark` in the connected-component matrix

5) Increment the marker for another component

6) Deal with edge cases on the fly

####2) `conBoundaries(x,N=4)` Detects internal and external boundaries of connected components in a logical matrix
#####Inputs

1) A logical matrix whose connected components are to be labeled

2) An integer to specify 4 or 8 connectivity

#####Output

Matrix with non-zero elements for boundaries of connected components. Each components are separately labeled.

#####Algorithm

Our algorithm detects boundaries by looking at neighboring elements (up, down, right and left). If there's a difference in value between the element and one of it's neighbors, it is determined to be a boundary point. It detects boundaries of components individually by first extracting only the component of interest. When performing row-major and column-major search, to increase the efficiency, we set the limit of the search area to the smallest rectangle that bounds the component we're looking at.

#####Steps

1) Matrix with connected components detected and separately labeled is obtained by passing the arguments to `conLabel`.

2) A matrix is initialized to store the boundaries of all the components.

3) A matrix is initialized to store the boundaries of each component separately.

4) Subset the component we are looking at. Extract the bounding rows and columns of that component.

5) For each component
* Perform a row-major search for boundary points.
+ + For edge cases (start and end point for the bound), check if they're non-zero. If non-zero mark as boundary and move on.
+ + Starting with the first element, check if the next column is different. If different, mark the non-zero element as boundary and move on.


+ Perform a column-major search for boundary points.
+  + For edge cases (start and end point for the bound), check if they're non-zero. If non-zero mark as boundary.
+  + Starting with the first element, check if the next row is different. If different, mark the non-zero element as boundary and move on.

* Add the boundaries of current component to the matrix storing boundaries for all components and move on to the next component.



####3) `imgMat(x)` Converts a PNG file to a logical matrix

#####Input

Image or image path

#####Output

A logical matrix indicating white or black pixels of the image

#####Algorithm

Here we convert a .png image file to a logical matrix to be further used as an input to conLabel and conBoundaries. 0 indicates white (and slightly offwhite) pixels and 1 indicates colored pixels of a high-contrast image.

#####Steps

1) Read the image using command `readPNG` from package `png`

2) Sum up the layers of the 3-D matrix obtained and normalize the data to be between 0 and 1 such that 0 indicates white and black indicates 1

3) Set all values greater then 0.1 to 1 and less than or equal to 0.1 to 0

We then build the final R package with these 3 functions using the package `devtools`. 


## Some examples
We ran our functions to check for boundaries for a logical matrix we created with a few components as well as holes in them and also for a sample image. The resulting images for both the examples are shown below.

### Example 1: Sample Matrix
We tested our functions with a 100x100 matrix defined below.
```{r}
library(Matrix)
library(concomp)
mat = matrix(rep(0, 100*100), nrow = 100)
mat[1:5, 1:100] = 1
mat[1:100, 1:5] = 1
mat[96:100, 1:100] = 1
mat[1:100, 96:100] = 1
mat[11:15, 11:90] = 1
mat[11:90, 11:15] = 1
mat[86:90, 11:90] = 1
mat[11:90, 86:90] = 1
mat[25:50, 25:50] = 1
mat[51:75,51:75] = 1
```

We tested the functions for both 4 and 8-connected connectivity. The results are shown below, we can clearly see that the two square blocks on the center are connected according to 8-connected connectivity but not according to 4-connected.

<p style='text-align: center;'> __Plot showing original matrix__</p>
  ```{r, fig.align='center'}
image(Matrix(mat))
```


<p style='text-align: center;'> __Plot showing different 4-connected components as detected by `conLabel()` function__</p>
  ```{r, echo=FALSE, fig.align='center'}
A = conLabel(mat)
image(Matrix(A), col.regions = c(1:max(A)))
```

<p style='text-align: center;'> __Plot showing different 8-connected components as detected by `conLabel()` function__</p>
  ```{r, echo=FALSE, fig.align='center'}
A = conLabel(mat,8)
image(Matrix(A), col.regions = c(1:max(A)))
```

<p style='text-align: center;'> __Plot showing boundaries of different 4-connected components as detected by `conLabel()` function__</p>
  ```{r, echo=FALSE, fig.align='center'}
A = conBoundaries(mat)
image(Matrix(A), col.regions = c(1:max(A)))
```


### Example 2: Sample png image
```{r, fig.align='center', echo=FALSE}
img = readPNG("git.png")
mat = imgMat("git.png")
plot(c(0,1000),c(0,400), axes=F, type ="n", xlab ="", ylab ="", main = "Original Image")
rasterImage(img, 0,0,1000,400, asp = .4)
```

<p style='text-align: center;'> __Matrix representation after changing the image into logical matrix using `imgMat()` function__</p>
  ```{r, fig.align='center', echo=FALSE}
mat = imgMat("git.png")
Matrix::image(Matrix(mat), col.regions = c(1:max(mat)))
```

<p style='text-align: center;'> __Plot showing different components as detected by `conLabel()` function__</p>
  ```{r, fig.align='center', echo=FALSE}
lab = conLabel(mat)
Matrix::image(Matrix(lab), col.regions = c(1:max(lab)))
```

<p style='text-align: center;'> __Plot showing boundaries of different components as detected by `conBoundaries()` function__</p>
  ```{r, fig.align='center', echo=FALSE}
labBound = conBoundaries(mat)
Matrix::image(Matrix(labBound), col.regions = c(1:max(labBound)))
```

