---
title: "STA 523 Final Project"
author: "Dipesh Gautam, Radhika Anand, Haosheng Luo, Siyang Li"
date: "12/12/2014"
output: html_document
---

All codes can be found at `https://github.com/ra125/concomp`.

For this project, we have built a package called `concomp` to detect connected components and holes and their boundaries in an image or a logical matrix.

##Motivation
While working on 523 Homework-3, where we had to detect Manhattan precinct boundaries, we had data points in all of Manhattan apart from the Central Park area. While trying to detect the Central Park, we were looking for functions to detect holes and boundaries and hence decided to create a package with functions for the same.

##Functions
We have written the following three functions in our package:

1) `conLabel()` Assigns labels to connected components in a logical matrix

2) `conBoundaries()` Detects internal and external boundaries of connected components in a logical matrix

3) `imgMat()` Converts a PNG file to a logical matrix

##Methodology

####1) `conLabel(x,N=4)` Assigns labels to connected components in a logical matrix

#####Inputs

1) A logical matrix whose connected components are to be labeled

2) An integer to specify 4 or 8 connectivity

#####Output

Matrix with connected components detected and separately labeled

#####Algorithm

We have implemented a single-pass algorithm where we traverse the matrix row wise and label the connected components using 4 or 8-connectivity as specified by the user. In 4-connectivity, we look at the top, left, right and bottom neighbor of each cell and in 8-connectivity we check all 8 neighbors.

#####Steps

1) A connected-component matrix is initialized to size of the input matrix

2) A `mark` is initialized and incremented for every detected component in the matrix

3) A row-major scan is started for the entire matrix

4) If a component cell is detected, the following steps are repeated while (`index` !=0) 
+ Set the corresponding cell to 0 in the matrix
+ A vector `index` is updated with all the neighboring cells (4 or 8) of the currently set cells 
+ Only the unique cells are retained 
+ Set the cells indicated by `index` to `mark` in the connected-component matrix

5) Increment the marker for another component

6) Deal with edge cases on the fly

####2) `conBoundaries(x,N=4)` Detects internal and external boundaries of connected components in a logical matrix
#####Inputs

1) A logical matrix in which boundaries are to be detected

2) An integer to specify 4 or 8 connectivity

#####Output

Matrix with non-zero elements for boundaries of connected components. Boundary of each component is separately labeled.

#####Algorithm

Our algorithm detects boundaries by looking at neighboring elements (up, down, right and left). If there's a difference in value between the element and one of it's neighbors, it is determined to be a boundary point. It detects boundaries of components individually by first extracting only the component of interest. When performing row-major and column-major search, to increase the efficiency, we set the limit of the search area to the smallest rectangle that bounds the component we're looking at.

#####Steps

1) Matrix with connected components detected and separately labeled is obtained by passing the arguments to `conLabel`.

2) A matrix is initialized to store the boundaries of all the components.

3) A matrix is initialized to store the boundaries of each component separately.

4) Subset the component we are looking at. Extract the bounding rows and columns of that component.

5) For each component
* Perform a row-major search for boundary points.
+ + For edge cases (start and end point for the bound), check if they're non-zero. If non-zero mark as boundary and move on.
+ + Starting with the first element, check if the next column is different. If different, mark the non-zero element as boundary and move on.


+ Perform a column-major search for boundary points.
+  + For edge cases (start and end point for the bound), check if they're non-zero. If non-zero mark as boundary.
+  + Starting with the first element, check if the next row is different. If different, mark the non-zero element as boundary and move on.

* Add the boundaries of current component to the matrix storing boundaries for all components and move on to the next component.


####3) `imgMat(x)` Converts a PNG file to a logical matrix

#####Input

Image or image path

#####Output

A logical matrix indicating white or black pixels of the image

#####Algorithm

Here we convert a .png image file to a logical matrix to be further used as an input to conLabel and conBoundaries. 0 indicates white (and slightly offwhite) pixels and 1 indicates colored pixels of a high-contrast image.

#####Steps

1) Read the image using command `readPNG` from package `png`

2) Sum up the layers of the 3-D matrix obtained and normalize the data to be between 0 and 1 such that 0 indicates white and black indicates 1

3) Set all values greater then 0.1 to 1 and less than or equal to 0.1 to 0

##Package Creation
Another motivation for this project was to have hands-on experience with package creation in R. Since, this was our first attemt at creating a package we only went through the minimum requirements of a package:

 * `R/` folder containing the R files
 + `man/` folder containing function documentation
 + `DESCRIPTION` file with package metadata

We used package `devtools` to build and install the package. Since our package uses a function from `png` package, and R's builtin `install.packages` function was not correctly installing the dependent packages, we used `install` function from `devtools` package to test the installation and validity of the package. We see that the package works as expected.

## Some examples
We ran our functions to check for boundaries for a logical matrix we created with a few components as well as holes in them and also for a sample image. The resulting images for both the examples are shown below.

To display the results of our functions while avoiding errors that might arise while installing the function due to dependence on the `png` package, we have copied the definitions for all functions and also sourced the packages required for testing, here. But if we install our package `concomp` correctly using `devtools::install("concomp")`, then all functions work directly from there. 

```{r,echo=FALSE}
check_packages = function(names)
{
  for(name in names)
  {
    if (!(name %in% installed.packages()))
      install.packages(name, repos="http://cran.us.r-project.org")
    
    library(name, character.only=TRUE, quietly=TRUE, warn.conflicts=FALSE, 
            verbose=FALSE)
  }
}
conLabel = function(mat, N=4)
{
  #Employ single-pass algorithm
  #Check whether the input is valid.
  #Components are either 4-connected or 8-connected.
  
  stopifnot(is.matrix(mat))
  stopifnot(sum(mat*(1-mat))==0)
  stopifnot(N==4||N==8)
  
  m=nrow(mat)
  n=ncol(mat)
  
  #Set up a zeros matrix for saving labels later.
  connec=matrix(rep(0,m*n),nrow=m,ncol=n)
  
  #Initialize the region label. 
  mark=1
  
  #Search by each row and column.
  for(i in 1:m)
  {
    for(j in 1:n)
    {
      if(mat[i,j]==1)
      {
        #Position of the first detected element in the region.
        index=(j-1)*m+i
        
        #Label the first element.
        connec[index]=mark
        while(length(index)!=0)
        {
          #Set visited element equal to zero, that we will not visit again.
          mat[index]=0
          neighbors=list()
          for(k in 1:length(index))
          {
            #If the element is in the bottom of the mat, it will only have left, right, and upper neighbors.
            if(index[k]%%m==0 && index[k]!=m*n)
            {
              if (N==4){
                offset=c(-1,m,-m)
              }else{
                offset=c(-1,m,-m, m-1,-m-1)
              }
              #If the element is in the top of the mat, it will only have left, right, and lower neighbors.
            } else if(index[k]%%m==1 && index[k]!=(n-1)*m+1)
            {
              if (N==4){
                offset = c(1,m,-m)
              }else{
                offset = c(1,m,-m, m+1, -m+1)
              }
              #If the element is in the very right of the mat, it will only have left, upper and lower neighbors.
            } else if(index[k]%/%m==n-1 && index[k]!=(n-1)*m+1 && index[k]!=m*n)
            {
              if(N==4){
                offset=c(1,-1,-m)
              }else{
                offset=c(1,-1,-m, -m+1,-m-1)
              }
              #If the element is in the bottom right corner of the mat, it will only have left and upper neighbors.
            } else if(index[k]==m*n)
            {
              if(N==4){
                offset=c(-1,-m)
              }else{
                offset= c(-1,-m, -m-1)
              }
              #If the element is in the topright corner of the mat, it will only have left and lower neighbors.
            } else if(index[k]==(n-1)*m+1)
            {
              if(N==4){
                offset=c(1,-m)
              }else{
                offset = c(1,-m,-m+1)
              }
              #Otherwise, the element has all four neighbors.
            } else {
              if(N==4){
                offset=c(-1,m,1,-m)
              }else{
                offset = c(-1,m,1,-m, m+1, m-1, -m-1, -m+1)
              }
            }
            
            neighbors=unlist(c(neighbors,index[k]+offset))
          }
          neighbors=unique(neighbors)
          
          #Get rid of all the neighbors that are out of the boundaries.
          neighbors=neighbors[neighbors>0]
          
          #Extract the position of all the neighbor elements that are in the region and label the element. 
          index=neighbors[which(mat[neighbors]==1)]
          connec[index]=mark
        }
        
        #Update the label before moving to the next region.
        mark=mark+1
      }
    }
  }
  
  return(connec)
  
}

conBoundaries = function(bw,N=4)
{
  #Check whether the input is valid.
  #Components are either 4-connected or 8-connected.
  
  stopifnot(is.matrix(bw))
  stopifnot(sum(bw*(1-bw))==0)
  stopifnot(N==4||N==8)
  
  #Label the regions in the image.
  L = conLabel(bw,N)
  num_labels = max(L)
  
  #Set up a zeros matrix for saving boundaries of the whole image later.
  BOUNDARY = L
  BOUNDARY[,]=0
  
  #Find boundaries for each region.
  for (n in 1:num_labels){
    segment = L
    #Set up a zeros matrix for saving boundaries of an individual region.
    boundary = L
    boundary[,]=0
    #Subset the region we are looking at. Extract the maxmimum and minimum rows and columns of that region.
    segment[which(!segment==n)]=0
    all.points = which(segment==n, arr.ind=T)
    row.limit = c(min(all.points[,1]), max(all.points[,1]))
    col.limit = c(min(all.points[,2]), max(all.points[,2]))
    
    #For each row, search the positions where the elements change from zero to the label or vice versa.
    for (i in row.limit[1]:row.limit[2]){
      for(j in col.limit[1]:col.limit[2]){
        #When we are in the first column of the image, and if the element is in the region, this is the boundary point. 
        if(j==1 && segment[i,j]==n){
          boundary[i,j] = n
        } else if(j==1 && segment[i,j]!=n){
          boundary[i,j] = 0
        #When we are in the last column of the image, and if the element is in the region, this is the boundary point.
        } else if(j==ncol(segment)&&segment[i,j]==n){
          boundary[i,j] = n
        } else if(j==ncol(segment)&&segment[i,j]!=n){
          boundary[i,j] = 0    
        #When the value of two adjacent elements are different, the bigger one is the boundary point (the background is labeled with zero).
        } else if(segment[i,j]>segment[i,j-1]){
          boundary[i,j] = n
        } else if (segment[i,j]>segment[i,j+1]){
          boundary[i,j] = n
        }
      }
      
    }
    #For each column, search the positions where the elements change from zero to the label or vice versa.
    for (j in col.limit[1]:col.limit[2]){
      for(i in row.limit[1]:row.limit[2]){
        #When we are in the first row of the image, and if the element is in the region, this is the boundary point. 
        if(i==1 && segment[i,j]==n){
          boundary[i,j] = n
        } else if(i==1 && segment[i,j]!=n){
          boundary[i,j] = 0
        #When we are in the last row of the image, and if the element is in the region, this is the boundary point.
        } else if(i==nrow(segment) && segment[i,j]==n){
          boundary[i,j] = n
        } else if(i==nrow(segment) && segment[i,j]!=n){
          boundary[i,j] = 0
        #When the value of two adjacent elements are different, the bigger one is the boundary point (the background is labeled with zero).
        } else if(segment[i,j]>segment[i-1,j]){
          boundary[i,j] = n
        } else if (segment[i,j]>segment[i+1,j]){
          boundary[i,j] = n
        }
      } 
    }   
    #Add all the individual region boundaries.
    BOUNDARY = BOUNDARY+boundary
  }
  return(BOUNDARY)
}

imgMat = function(png)
{
  #Read the PNG file
  img = readPNG(png)
  
  #Sum up all the layers and normalize the image data
  scl_img=img[,,1]+img[,,2]+img[,,3]
  scl_img=scl_img/max(scl_img)
  
  #Set white pixels equal to 0 and black to 1
  scl_img=1-scl_img
  scl_img[which(scl_img>0.1)]=1
  scl_img[which(scl_img<=0.1)]=0
  return(scl_img)
}
```
### Example 1: Sample Matrix

We tested our functions with a 100x100 matrix defined below.
```{r}
suppressMessages(check_packages(c('Matrix', 'png', 'graphics')))
mat = matrix(rep(0, 100*100), nrow = 100)
mat[1:5, 1:100] = 1
mat[1:100, 1:5] = 1
mat[96:100, 1:100] = 1
mat[1:100, 96:100] = 1
mat[11:15, 11:90] = 1
mat[11:90, 11:15] = 1
mat[86:90, 11:90] = 1
mat[11:90, 86:90] = 1
mat[25:50, 25:50] = 1
mat[51:75,51:75] = 1
```

We tested the functions for both 4 and 8-connected connectivity. The results are shown below, we can clearly see that the two square blocks on the center are connected according to 8-connected connectivity but not according to 4-connected.

<p style='text-align: center;'> __Plot showing original matrix__</p>
  ```{r, fig.align='center',echo=FALSE}
image(Matrix(mat))
```


<p style='text-align: center;'> __Plot showing different 4-connected components as detected by `conLabel()` function__</p>
  ```{r, echo=FALSE, fig.align='center'}
A = conLabel(mat)
image(Matrix(A), col.regions = c(1:max(A)))
```

<p style='text-align: center;'> __Plot showing different 8-connected components as detected by `conLabel()` function__</p>
  ```{r, echo=FALSE, fig.align='center'}
A = conLabel(mat,8)
image(Matrix(A), col.regions = c(1:max(A)))
```

<p style='text-align: center;'> __Plot showing boundaries of different 4-connected components as detected by `conLabel()` function__</p>
  ```{r, echo=FALSE, fig.align='center'}
A = conBoundaries(mat)
image(Matrix(A), col.regions = c(1:max(A)))
```


### Example 2: Sample png image
```{r, fig.align='center', echo=FALSE}
img = readPNG("~dg156/finalData/git.png")
mat = imgMat("~dg156/finalData/git.png")
plot(c(0,1000),c(0,400), axes=F, type ="n", xlab ="", ylab ="", main = "Original Image")
rasterImage(img, 0,0,1000,400, asp = .4)
```

<p style='text-align: center;'> __Matrix representation after changing the image into logical matrix using `imgMat()` function__</p>
  ```{r, fig.align='center', echo=FALSE}
mat = imgMat("~dg156/finalData/git.png")
Matrix::image(Matrix(mat), col.regions = c(1:max(mat)))
```

<p style='text-align: center;'> __Plot showing different components as detected by `conLabel()` function__</p>
  ```{r, fig.align='center', echo=FALSE}
lab = conLabel(mat)
Matrix::image(Matrix(lab), col.regions = c(1:max(lab)))
```

<p style='text-align: center;'> __Plot showing boundaries of different components as detected by `conBoundaries()` function__</p>
  ```{r, fig.align='center', echo=FALSE}
labBound = conBoundaries(mat)
Matrix::image(Matrix(labBound), col.regions = c(1:max(labBound)))
```

