---
title: "STA 523 Final Project"
author: "Dipesh Gautam, Radhika Anand, Haosheng Luo, Siyang Li"
date: "12/12/2014"
output: html_document
---

For this project, we have built a package called `concomp` to detect connected components and holes and their boundaries in an image or a logical matrix.

##Motivation
While working on 523 Homework-3, where we had to detect Manhattan precinct boundaries, we had data points in all of Manhattan apart from the Central Park area. While trying to detect the Central Park, we were looking for functions to detect holes and boundaries and hence decided to create a package with functions for the same.

##Functions
We have written the following three functions in our package:

1) `conLabel()` Assigns labels to connected components in a logical matrix

2) `conBoundaries()` Detects internal and external boundaries of connected components in a logical matrix

3) `imgMat()` Converts a PNG file to a logical matrix

##Methodology

#####1) `conLabel(x,N=4)` Assigns labels to connected components in a logical matrix

#####Inputs

1) A logical matrix whose connected components are to be labeled

2) An integer to specify 4 or 8 connectivity

#####Output

Matrix with connected components detected and separately labeled

#####Algorithm

We have implemented a single-pass algorithm where we traverse the matrix row wise and label the connected components using 4 or 8-connectivity as specified by the user. In 4-connectivity, we look at the top, left, right and bottom neighbor of each cell to check whether they belong to the same component and in 8-connectivity we check all 8 neighbors.

#####Steps

1) A connected-component matrix is initialized to size of the input matrix

2) A mark is initialized and incremented for every detected object in the image.

3) A row-major scan is started for the entire image.

4) If an object pixel is detected, then following steps are repeated while (Index !=0) 
+ Set the corresponding pixel to 0 in Image. 
+ A vector (Index) is updated with all the neighboring pixels of the currently set pixels. 
+ Unique pixels are retained and repeated pixels are removed. 
+ Set the pixels indicated by Index to mark in the connected-component matrix.

5) Increment the marker for another object in the image.

2) `conBoundaries(x,N=4)` Detects internal and external boundaries of connected components in a logical matrix



3) `imgMat(x)` Converts a PNG file to a logical matrix



We then built the R package using the package `devtools`. 


---
title: "STA523 Final Project"
author: "Dipesh Gautam, Radhika Anand, Haosheng Luo, Siyang Li"
date: "12/12/2014"
output: html_document
---


```{r}
library(Matrix)
library(concomp)

```


## Some examples
We ran our functions to check for boundaries for a logical matrix we created with a few components as well as holes in them and also for a sample image. The resulting images for both the examples are shown below.

### Example 1: Sample Matrix
We tested our functions with a 100x100 matrix defined below.
```{r}
mat = matrix(rep(0, 100*100), nrow = 100)
mat[1:5, 1:100] = 1
mat[1:100, 1:5] = 1
mat[96:100, 1:100] = 1
mat[1:100, 96:100] = 1
mat[11:15, 11:90] = 1
mat[11:90, 11:15] = 1
mat[86:90, 11:90] = 1
mat[11:90, 86:90] = 1
mat[25:50, 25:50] = 1
mat[51:75,51:75] = 1
```

We tested the functions for both 4 and 8-connected connectivity. The results are shown below, we can clearly see that the two square blocks on the center are connected according to 8-connected connectivity but not according to 4-connected.

<p style='text-align: center;'> __Plot showing original matrix__</p>
```{r, fig.align='center'}
image(Matrix(mat))
```


<p style='text-align: center;'> __Plot showing different 4-connected components as detected by `conLabel()` function__</p>
```{r, echo=FALSE, fig.align='center'}
A = conLabel(mat)
image(Matrix(A), col.regions = c(1:max(A)))
```

<p style='text-align: center;'> __Plot showing different 8-connected components as detected by `conLabel()` function__</p>
```{r, echo=FALSE, fig.align='center'}
A = conLabel(mat,8)
image(Matrix(A), col.regions = c(1:max(A)))
```

<p style='text-align: center;'> __Plot showing boundaries of different 4-connected components as detected by `conLabel()` function__</p>
```{r, echo=FALSE, fig.align='center'}
A = conBoundaries(mat)
image(Matrix(A), col.regions = c(1:max(A)))
```


### Example 2: Sample png image
```{r, fig.align='center', echo=FALSE}
img = readPNG("git.png")
mat = imgMat("git.png")
plot(c(0,1000),c(0,400), axes=F, type ="n", xlab ="", ylab ="", main = "Original Image")
rasterImage(img, 0,0,1000,400, asp = .4)
```

<p style='text-align: center;'> __Matrix representation after changing the image into logical matrix using `imgMat()` function__</p>
```{r, fig.align='center', echo=FALSE}
mat = imgMat("git.png")
Matrix::image(Matrix(mat), col.regions = c(1:max(mat)))
```

<p style='text-align: center;'> __Plot showing different components as detected by `conLabel()` function__</p>
```{r, fig.align='center', echo=FALSE}
lab = conLabel(mat)
Matrix::image(Matrix(lab), col.regions = c(1:max(lab)))
```

<p style='text-align: center;'> __Plot showing boundaries of different components as detected by `conBoundaries()` function__</p>
```{r, fig.align='center', echo=FALSE}
labBound = conBoundaries(mat)
Matrix::image(Matrix(labBound), col.regions = c(1:max(labBound)))
```

